Sampling Strategies in Decisions from Experience

### 2022-05-21 manuscript.Rmd
Update Section 3: Expectations

The model assumes that such comparison rounds are repeatedly carried out until the cumulative sum of round wins for one of the two prospects reaches a threshold $\theta \in \mathbb{N}$.
Similar to the switching probability, it is assumed that the threshold $\theta$ is adopted prior the start of the sampling and accumulation process and is fixed throughout.
For $P^X$, then, one obtains a sequence of mean comparisons, denoted as $Z^X_1, Z^X_2, ...$, and an evidence accumulation process taking the form of a random walk $\{D^X_v\}_{v \in \mathbb{N}}$ defined by
$$
D^X_v = \sum_{c = 1}^v Z^X_c, \quad c,v \in \mathbb{N}
\;.
\tag{3}
$$
$D^X_v$ is the cumulative sum of round wins of prospect $P^X$ after $v$ comparison rounds, and resembles a dynamic decision variable.
Note that since (2) describes a Bernoulli trial, the random walk in (3) is based on a Bernoulli process.
Prospect $P^X$ would be chosen, if $D^Y_v < \theta \leq D^X_v$ is satisfied, and vice versa.

@@

To implement the evidence accumulation process, the summary integration model assumes that an agent continues sampling from both prospects until the cumulative sum across all outcomes sampled from one of the prospects reaches a threshold $\theta$. 
As in the round-wise integration model, the threshold $\theta$ is assumed to be adopted prior the start of the sampling and accumulation process and is fixed throughout.
For $P^X$, then, one obtains a sequence of sampled outcomes, $X_1, X_2, ...$, and an evidence accumulation process taking the form of a random walk $\{D^X_t\}_{t \in \mathbb{N}}$ defined by 
$$
D^X_t = \sum_{k = 1}^t X_k, \quad k,t \in \mathbb{N}
\; ,
\tag{4}
$$
where $D^X_t$ is the cumulative sum after $t$ outcomes sampled from prospect $P^X$ and resembles a dynamic decision variable.
Assuming that $D^Y_m$ is the cumulative sum after $m$ sampled outcomes from prospect $P^Y$, prospect $P^X$ would be chosen, if $D^Y_m < \theta \leq D^X_t$ is satisfied, and vice versa.

Note again that the described model assumes two random walks---i.e., one for each prospect---approaching the same threshold, which now is the cumulative sum across all sampled outcomes a prospect must produce in order to be chosen.
However, one may also define a single random walk, resembling a common dynamic decision variable, which approaches a positive threshold $\theta^+$ and a negative threshold $\theta^-$.
That is, (4) may be substituted by
$$
D^X_{t,m} = \sum_{k = 1}^t X_k - \sum_{j = 1}^m Y_j \quad k,t,j,m \in \mathbb{N}
\; ,
\tag{5}
$$
where $D^X_{t,m}$ is the difference between the cumulative sums after $t$ and $m$ outcomes sampled from prospects $P^X$ and $P^Y$, respectively.
The thresholds thus determine how much larger the cumulative sum of a prospect must be in order for the prospect to be chosen.

@@

Below, the key predictions concerning the effects of the model parameters $\psi$ and $\theta$ on the length of the sampling and accumulation process and the resulting choice are outlined.
Moreover, previous research has shown that differences as to how information about the properties of prospects is processed can translate to characteristic shapes of cumulative prospect theory's (CPT) value and weighting function [see @pachurHowTwainCan2017; @zilkerNonlinearProbabilityWeighting2021].
Respectively, this section also contains predictions about the relations between $\psi$ and $\theta$ and the CPT parameters.
A summary of the model predictions can be found in Table\ \@ref(tab:predictions).

(ref:predictions) Summary of Model Predictions

```{r predictions}
predictions <- tibble("Measure" = 
                        c("Choices pattern",
                          "",
                          "$\\gamma$ (curvature of $w$)",
                          "$\\delta$ (elevation of $w$)",
                          "$\\alpha$ (concavity of $v$)"),
                      "Predictions for changes in $\\psi$" = 
                        c("More EV maximization for decreasing $\\psi$",
                          "More as-if underweighting of rare outcomes for increasing $\\psi$",
                          "$\\geq 1$, increases with $\\psi$",
                          "1, no effect of $\\psi$",
                          "$\\leq 1$, decreases with increasing $\\psi$"))
apa_table(predictions, caption = "(ref:predictions)", note = "The predictions should become more robust for high values of $\\theta$.", escape = FALSE)
```

## Length of Information-Processing Chains

@@

## Final Choice

@@

Before elaborating how the above model predictions may translate into estimates of the CPT parameters, the following section reviews how CPT in general and its weighting function in particular capture deviations from EV maximization.

@@

The choices people make between prospects are often *described* in terms of deviations from EV maximization, according to which the prospect with the largest EV
$$
EV = \sum_i^n x_i \times p_X(x_i)
\tag{10}
$$ 
should be chosen.
To describe how people's choices deviate from EV maximization [see @erevAnomaliesForecastsDescriptive2017, for a recent replication of classical demonstrations],^[<!--Open footnote-->
@tverskyAdvancesProspectTheory1992 addressed the *fourfold pattern of risk attitudes*---i.e., choices between prospects indicate risk aversion for gains of high probability and losses of low probability, and risk seeking for gains of low probability and losses of high probability---and related effects, e.g., the *certainty effect* and the *reflection effect*, where the latter were already addressed by CPT's predecessor, prospect theory [@kahnemanProspectTheoryAnalysis1979]. 
Together, these empirical choice patterns indicate a violation of EV and expected utility (EU) maximization [see @bernoulliExpositionNewTheory1954\/1738].
<!--Close footnote-->]
CPT and similar rank-dependent models [see @stottCumulativeProspectTheory2006, for an overview of models] fit choice data by assuming that people maximize the valuation
$$
V = \sum_i^n v(x_i) \times \pi_i
\; ,
\tag{11}
$$ 
with objective outcomes $x_i$ being transformed by a value function $v$, and *cumulative decision weights* $\pi_i$ being determined by the difference between transformed cumulative probabilities of the distribution $P'$.
More specifically, following @tverskyAdvancesProspectTheory1992, the objective outcomes are transformed by a value function
$$
\begin{aligned}
  v : \Omega' &\to \mathbb{R} \\
  x &\mapsto 
  \begin{cases}
     x_i^\alpha &\forall x_i \geq 0, \\
     -\lambda |x_i|^\alpha &\text{else}
     \; ,
  \end{cases}
\end{aligned}
\tag{12}
$$
with $\alpha \in [0,1]$ determining the degree of the function's concavity (convexity) over the positive (negative) outcome interval, and $\lambda > 1$ increasing the function's slope over the negative outcome interval only.
Each subjective value $v(x)$ is then multiplied (or: weighted) with a cumulative decision weight that takes the form
$$
\pi_i =
  \begin{cases}
     w^+(P(X \geq x_i)) - w^+(P(X > x_i)) \quad \forall x_i \geq 0, \\
     w^-(P(X \leq x_i)) - w^-(P(X < x_i)) \quad \text{else}
     \; , 
  \end{cases}
\tag{13}
$$
where $w$ is a monotonic increasing, nonlinear weighting function $w: [0,1] \to [0,1]$ satisfying $w^+(0) = w^-(0) = 0$ and $w^+(1) = w^-(1) = 1$.

Essentially, the model is such that the cumulative decision weights derived for the outcomes may be greater or smaller than their objective probabilities, causing the transformed outcomes---i.e., the subjective values---to be over- or underweighted *in* CPT, respectively.
Accordingly, for the description of choices, one may adopt an as-if weighting terminology by stating that people choose as if they maximized the value in (11) and applied the weighting pattern that was estimated in CPT.
Importantly, the as-if prefix indicates that it cannot be concluded from the estimated weighting pattern that the mind indeed performs any of the computations associated with the weighting of subjective values; 
rather, the mind processes the information about the properties of choice alternatives in a way that the resulting choices translate to the estimated weighting pattern [@gigerenzerHowExplainBehavior2020].
Respectively, this paper treats CPT as a data model and discusses---and refines in a simulation---what weighting patterns can be expected, were the mind to carry out the information-processing chains that are assumed by the proposed sampling and accumulation model [see @pachurHowTwainCan2017; @zilkerNonlinearProbabilityWeighting2021, for similar theoretical inquiries].

Now, several consequences for the weighting patterns that can be estimated in CPT follow from (13), which are briefly reviewed next:

* From the transformation of cumulative probabilities it follows that the value of a cumulative decision weight $\pi$ depends on the rank of the outcome $x$ for which $\pi$ is determined.

* The value of $\pi$ depends on the estimated shape of the graph of $w$, i.e., the weighting function's parameters.
Specifically, the shape of the graph of $w$ displays over which interval on the cumulative probability scale $[0,1]$ the images of $w$ take values that are greater or smaller than the respective cumulative probabilities and how much greater or smaller these images are.
Hence, for a nonlinear graphical shape of $w$, the same numerical difference between objective cumulative probabilities may translate to cumulative decision weights of different value.

* For prospects containing either only positive or only negative outcomes, $\sum_i^n\pi_i = 1$ is satisfied.
Accordingly, for such prospects, the weighting of subjective values with cumulative decision weights rather than with objective probabilities may be roughly understood as a redistribution of the entire probability mass of $P(\Omega') = 1$ across outcomes [@zilkerNonlinearProbabilityWeighting2021].

\noindent
Because the value of each $\pi$ depends on the two transformed cumulative probabilities in (13)---i.e., the probability of obtaining a positive (negative) outcome equal to or greater (smaller) than a respective outcome $x$, and the probability of obtaining a strictly greater (smaller) outcome---the remaining consequences for the possible weighting patterns are reviewed by considering the actual weighting function.
Therefore, Figure\ \@ref(fig:weighting-function) illustrates some of the possible graphical shapes of the two-parameter weighting function of @goldsteinExpressionTheoryPreference1987, which, however, is just one of several parameterizations that have been proposed [e.g., @prelecProbabilityWeightingFunction1998; @tverskyAdvancesProspectTheory1992; see @stottCumulativeProspectTheory2006, for an overview].
Now, let each $p$ on the abscissa be one of the four cumulative probabilities from (13). 
Then each graph in Figure\ \@ref(fig:weighting-function) displays the graphical shape of the weighting function
$$
\begin{aligned}
  w : [0,1] &\to [0,1] \\
  p &\mapsto \frac{\delta \times p^{\gamma}}
  {\delta \times p^{\gamma} + (1-p)^{\gamma}}
  \; ,
\end{aligned}
\tag{14}
$$
for the respective values of the parameters $\gamma \in [0,2]$ and $\delta > 0$.
Evidently, both parameters have distinct effects on the graphs' shape, with $\gamma$ affecting the curvature and $\delta$ the elevation.
As a consequence from (13), then, each combination of parameters implies a particular weighting pattern, with some of them being similar and others being rather distinct.

@@

#plot shapes of weighting function
wf %>% 
  ggplot(aes(p, wp, group = gamma)) +
  facet_wrap(~delta, labeller = as_labeller(label_delta, default = label_parsed)) + 
  scale_x_continuous(breaks = seq(0, 1, .5)) +
  scale_y_continuous(breaks = seq(0, 1, .5)) + 
  labs(x = expression(p),
       y = expression(w(p)), 
       color = expression(gamma)) + 
  theme_minimal() + 
  geom_line(aes(color = gamma), size = .5) +
  scale_color_viridis(option = "inferno") + 
  geom_abline(intercept = 0, slope = 1, color = "gray", linetype = "dashed")
```

@@

To simplify matters, the further discussion and the computational analysis are constrained to choices between a safe prospect---possessing just one outcome that occurs with certainty---and a two-outcome prospect (hereafter: risky prospect).
All outcomes are assumed to be positive and the safe outcome is assumed to fall between the the low-rank outcome, $x_{low}$, and the high-rank outcome, $x_{high}$, of the risky prospect to omit dominance.
A summary of the predictions for the CPT parameters can be found in Table\ \@ref(tab:predictions).

@@

@@

>>>>

The model assumes that such comparison rounds are repeatedly carried out until the cumulative sum of round wins for one of the two prospects reaches a threshold $\theta \in \mathbb{N}$.
Similar to the switching probability, it is assumed that the threshold $\theta$ is adopted prior the start of the sampling and accumulation process and is fixed throughout.
For $P^X$, then, one obtains a sequence of mean comparisons, denoted as $Z^X_1, Z^X_2, ...$, and an evidence accumulation process taking the form of a random walk $\{D^X_v\}_{v \in \mathbb{N}}$ defined by
$$
D^X_v = \sum_{c = 1}^v Z^X_c, \quad c,v \in \mathbb{N}
\;.
\tag{3}
$$
$D^X_v$ is the cumulative sum of round wins of prospect $P^X$ after $v$ comparison rounds, and resembles a dynamic decision variable.
Note that since (2) describes a Bernoulli trial, the random walk in (3) is based on a Bernoulli process.
Prospect $P^X$ would be chosen once $D^Y_v < \theta \leq D^X_v$ is satisfied, and vice versa.

@@

To implement the evidence accumulation process, the summary integration model assumes that an agent continues sampling from both prospects until the cumulative sum across all outcomes sampled from one of the prospects reaches a threshold $\theta$. 
As in the round-wise integration model, the threshold $\theta$ is assumed to be adopted prior the start of the sampling and accumulation process and is fixed throughout.
For $P^X$, then, one obtains a sequence of sampled outcomes, $X_1, X_2, ...$, and an evidence accumulation process taking the form of a random walk $\{D^X_t\}_{t \in \mathbb{N}}$ defined by 
$$
D^X_t = \sum_{k = 1}^t X_k, \quad k,t \in \mathbb{N}
\; ,
\tag{4}
$$
where $D^X_t$ is the cumulative sum across $t$ outcomes sampled from prospect $P^X$ and resembles a dynamic decision variable.
Assuming that $D^Y_m$ is the cumulative sum across $m$ sampled outcomes from prospect $P^Y$, prospect $P^X$ would be chosen once $D^Y_m < \theta \leq D^X_t$ is satisfied, and vice versa.

Note again that the described model assumes two random walks---i.e., one for each prospect---approaching the same threshold, which now is the cumulative sum across all sampled outcomes a prospect must produce in order to be chosen.
However, one may also define a single random walk, resembling a common dynamic decision variable, which approaches a positive threshold $\theta^+$ and a negative threshold $\theta^-$.
That is, (4) may be substituted by
$$
D^X_{t,m} = \sum_{k = 1}^t X_k - \sum_{j = 1}^m Y_j \quad k,t,j,m \in \mathbb{N}
\; ,
\tag{5}
$$
where $D^X_{t,m}$ is the difference between the cumulative sums across $t$ and $m$ outcomes sampled from prospects $P^X$ and $P^Y$, respectively.
The thresholds thus determine how much larger the cumulative sum of a prospect must be in order for the prospect to be chosen.

@@

Below, it is discussed which effects the key model parameters, i.e., the switching probability $\psi$ and the threshold $\theta$, could have on the choices and the corresponding as-if weighting pattern that are implied by the round-wise and the summary integration model.
The effects of the switching probability $\psi$ should be informative about the interplay between different sampling strategies and the integration/decision strategy each model resembles.
Assuming that both models aim to assess true differences in the latent properties of prospects, the effects of the threshold $\theta$ should indicate whether these differences are indeed more accurately assessed---thus, leading to more systematic choices and as-if weighting patterns---the more evidence is accumulated.
In addition, previous research has shown that differences as to how information about the properties of prospects is processed can translate to distinct signatures in CPT's psychoeconomic value and weighting function, without actually assuming these functions to be computed whatsoever [see @pachurHowTwainCan2017; @stewartDecisionSampling2006; @zilkerNonlinearProbabilityWeighting2021].^[<!--Open footnote-->
Note that it has been argued that the distinction between CPT's functions used to describe choice data and models aiming to explain how this data is generated in first place are reminiscent of Marr's [-@marrVisionComputationalInvestigation1982] different levels of explanation [@zilkerMeasuringModelingConstruction2020; also cf. @griffithsProbabilisticModelsCognition2010].
That is, whereas CPT may be considered a computational-level theory that describes an abstract solution to the problem the mind aims to solve when offered a choice between prospects, process models [cf. @jareckiFrameworkBuildingCognitive2020] such as heuristics and sequential sampling models may be considered algorithmic-level theories that specify how the mind can approximate the abstract solution with the capacity limits it possesses.
<!--Close footnote-->]
Accordingly, this section also formulates expectations regarding the potential relations between the switching probability $\psi$ and threshold $\theta$ (data-generating parameters) and the CPT parameters (data-modeling parameters).
Summaries of the expected model behaviors can be found in Tables\ \@ref(tab:expectations-roundwise) and \@ref(tab:expectations-summary).

## Length of Sampling and Accumulation Processes and Final Choice

### Round-Wise Integration Model

@@

@@

Before elaborating how the expectation outlined above may translate into estimates of the CPT parameters, the following section reviews how CPT in general and its weighting function in particular capture deviations from EV maximization.

@@

The actual choices people make between prospects are often *described* in terms of deviations from EV maximization, according to which the prospect with the largest EV
$$
EV = \sum_{i \in \mathbb{N}}^n x_i \times p_X(x_i)
\tag{6}
$$ 
should be chosen.
To describe how people's choices deviate from EV maximization [see @erevAnomaliesForecastsDescriptive2017, for a recent replication of classical demonstrations; see also @kahnemanProspectTheoryAnalysis1979; @tverskyAdvancesProspectTheory1992], CPT and similar rank-dependent models [see @stottCumulativeProspectTheory2006, for an overview of models] fit choice data by assuming that people maximize the valuation
$$
V = \sum_{i \in \mathbb{N}}^n v(x_i) \times \pi_i
\; ,
\tag{7}
$$ 
with objective outcomes $x_i$ being transformed by a value function $v$, and *cumulative decision weights* $\pi_i$ being determined by the difference between transformed cumulative probabilities of the distribution $P$.
More specifically, following @tverskyAdvancesProspectTheory1992, the objective outcomes are transformed by a value function
$$
\begin{aligned}
  v : \Omega' &\to \mathbb{R} \\
  x &\mapsto 
  \begin{cases}
     x_i^\alpha &\forall x_i \geq 0, \\
     -\lambda |x_i|^\alpha &\text{else}
     \; ,
  \end{cases}
\end{aligned}
\tag{8}
$$
with $\alpha \in [0,1]$ determining the degree of the function's concavity (convexity) over the positive (negative) outcome interval, and $\lambda > 1$ increasing the function's slope over the negative outcome interval only.
Each subjective value $v(x)$ is then multiplied (or: weighted) with a cumulative decision weight that takes the form
$$
\pi_i =
  \begin{cases}
     w^+(P(X \geq x_i)) - w^+(P(X > x_i)) \quad \forall x_i \geq 0, \\
     w^-(P(X \leq x_i)) - w^-(P(X < x_i)) \quad \text{else}
     \; , 
  \end{cases}
\tag{9}
$$
where $w$ is a monotonic increasing, nonlinear weighting function satisfying $w^+(0) = w^-(0) = 0$ and $w^+(1) = w^-(1) = 1$.
To simplify matters, the further discussion and the computational analysis are constrained to choices between a safe prospect---possessing just one outcome that occurs with certainty---and a two-outcome prospect (hereafter: risky prospect).
All outcomes are assumed to be positive and the safe outcome is assumed to fall between the the low-rank outcome, $x_{low}$, and the high-rank outcome, $x_{high}$, of the risky prospect to omit dominance.
One can then disregard the negative parts of (8) and (9), including the parameter $\lambda$ of the value function, and (9) simplifies to
$$
\begin{aligned}
  \pi_{high} &= w(p_x(x_{high})) \\
  \pi_{low} &= w(p_x(x_{low}) + p_x(x_{high})) - w(p_x(x_{high})) \\
  &= 1 - \pi_{high}
  \; .
\end{aligned}
\tag{10}
$$

Essentially, the model is such that the cumulative decision weights derived for the outcomes of the risky prospect may be greater or smaller than their objective probabilities, causing the transformed outcomes---i.e., the subjective values---to be over- or underweighted *within* CPT, respectively.
Accordingly, for the description of choices, one may adopt an as-if weighting terminology by stating that people choose as if they maximized the value in (7) and applied the weighting pattern that was estimated in CPT.
Importantly, the as-if prefix indicates that there is no claim that the mind indeed performs any of the computations associated with the weighting of subjective values; 
rather, the mind processes the information about the properties of choice alternatives in a way that the resulting choices translate to the estimated weighting pattern [cf. @gigerenzerHowExplainBehavior2020].
Accordingly, once again, this paper treats CPT as a data model and discusses---and refines in a simulation---what weighting patterns can be expected, were the mind to carry out the sampling and accumulation processes that are assumed by the presented round-wise and summary integration model [see @pachurHowTwainCan2017; @zilkerNonlinearProbabilityWeighting2021, for similar theoretical inquiries].

Now, several consequences for the weighting patterns that can be estimated in CPT follow from the exact specification of the weighting function $w$, which are briefly reviewed next:
Therefore, Figure\ \@ref(fig:weighting-function) illustrates some of the possible graphical shapes of the two-parameter weighting function of @goldsteinExpressionTheoryPreference1987, which, however, is just one of several parameterizations that have been proposed [e.g., @prelecProbabilityWeightingFunction1998; @tverskyAdvancesProspectTheory1992; see @stottCumulativeProspectTheory2006, for an overview].
Each graph in Figure\ \@ref(fig:weighting-function) displays the graphical shape of the weighting function
$$
\begin{aligned}
  w : [0,1] &\to [0,1] \\
  p_X(x_{high}) &\mapsto \frac{\delta \times (p_X(x_{high}))^{\gamma}}
  {\delta \times (p_X(x_{high}))^{\gamma} + (1-p_X(x_{high}))^{\gamma}}
  \; ,
\end{aligned}
\tag{11}
$$
for the respective values of the parameters $\gamma \in [0,2]$ and $\delta > 0$.
Evidently, both parameters have distinct effects on the graphs' shape, with $\gamma$ affecting the curvature and $\delta$ the elevation.
[As a consequence from (13), then, each combination of parameters implies a particular weighting pattern, with some of them being similar and others being rather distinct.]

@@

#plot shapes of weighting function
wf %>% 
  ggplot(aes(p, wp, group = gamma)) +
  facet_wrap(~delta, labeller = as_labeller(label_delta, default = label_parsed)) + 
  scale_x_continuous(breaks = seq(0, 1, .5)) +
  scale_y_continuous(breaks = seq(0, 1, .5)) + 
  labs(x = "Probability of High-Rank Risky Outcome",
       y = "Transformed Probability / Decision Weight", 
       color = expression(gamma)) + 
  theme_minimal() + 
  geom_line(aes(color = gamma), size = .5) +
  scale_color_viridis(option = "inferno") + 
  geom_abline(intercept = 0, slope = 1, color = "gray", linetype = "dashed")
```

@@

A summary of the predictions for the CPT parameters can be found in Table\ \@ref(tab:predictions).

@@

#### Round-Wise Integration Model

@@

(ref:expectations-roundwise) Expected Effects of Switching Probability $\psi$ and Threshold $\theta$ in the Round-Wise Integration Model

```{r expectations-roundwise}
exp_roundwise <- tibble("Measure" = 
                        c("Choices pattern",
                          "",
                          "$\\gamma$ (curvature of $w$)",
                          "$\\delta$ (elevation of $w$)",
                          "$\\alpha$ (concavity of $v$)"),
                      "Predictions for changes in $\\psi$" = 
                        c("More EV maximization for decreasing $\\psi$",
                          "More as-if underweighting of rare outcomes for increasing $\\psi$",
                          "$\\geq 1$, increases with $\\psi$",
                          "1, no effect of $\\psi$",
                          "$\\leq 1$, decreases with increasing $\\psi$"))

apa_table(exp_roundwise, caption = "(ref:expectations-roundwise)", note = "The predictions should become more robust for high values of $\\theta$.", escape = FALSE)
```

#### Summary Integration Model

(ref:expectations-summary) Expected Effects of Switching Probability $\psi$ and Threshold $\theta$ in the Summary Integration Model

```{r expectations-summary}
exp_summary <- tibble("Measure" = 
                        c("Choices pattern",
                          "",
                          "$\\gamma$ (curvature of $w$)",
                          "$\\delta$ (elevation of $w$)",
                          "$\\alpha$ (concavity of $v$)"),
                      "Predictions for changes in $\\psi$" = 
                        c("More EV maximization for decreasing $\\psi$",
                          "More as-if underweighting of rare outcomes for increasing $\\psi$",
                          "$\\geq 1$, increases with $\\psi$",
                          "1, no effect of $\\psi$",
                          "$\\leq 1$, decreases with increasing $\\psi$"))

apa_table(exp_summary, caption = "(ref:expectations-summary)", note = "The predictions should become more robust for high values of $\\theta$.", escape = FALSE)
```
